Experience Report: Haskell in Computational Biology

### 2012-02-09 mrfy_experience_report.tex
start of 'ugly code'

- profiling
-

>>>>

Despite the beauty of some of the code we were able to write, we found several 
instances where the easiest path seemed to be to write ugly code. The first of 
these relates to the otherwise elegant Viterbi function. Our stochastic search 
approach calls the Viterbi function a tremendous number of times, and most of 
the program's runtime is spent in this function. However, the Viterbi algorithm 
retains not just a score (the probability of the sequence given a state path) 
but also the state path itself. We implemented this state path as a list of 
\texttt{Int}, and every step in the recursion conses a new element onto the 
list. However, at each step of the stochastic search, we discard the state 
path; we are searching to optimize score only. We only need the state path at 
the termination of the search, when we wish the program to output an alignment 
of the query sequence to the Markov random field model. We found that 
optimizing the Viterbi function by removing this state path entirely improved 
runtime performance by nearly 50\%. However, we still needed the Viterbi 
function to provide a state path when we called it after search finally 
terminated. Thus, we wondered how to implement the Viterbi function so that a 
state path was optional.

In order to determine if this optimization was even worthwhile, we first wrote 
a quick and dirty approach that is shameful in any language, but easy: we 
duplicated code. We copied \texttt{viterbi} to produce a \texttt{viterbiF} 
which did not retain state path. Next, however, we wondered how to avoid code 
duplication. Clearly, littering the code with conditionals would produce code 
that was ugly as well as less efficient. We knew how we would solve this 
problem in other languages; we would have used macros in Lisp, or 
metaprogramming via \texttt{eval} in Ruby, to produce two versions of the 
function from one prototype. We struggled, however, to solve this problem 
idiomatically in Haskell.

breadcrumb: solution

Profiling: only top-level cost centers can be added automatically
 - had to pollute our code with cost center annotations
 - perhaps a better way to do this?

Awkwardness of trace and no backtraces led us to write ugly wrappers around Vector.slice and Vector.!

\section{Barriers}

- profiling
- debugging
